Reasoning About a Machine with Local Capabilities

### 2019-02-13 toplas/paper.tex
[TOPLAS] Consolidated POPL19 reference into one paragraph

\section{Discussion}
\label{sec:discussion}
\itoplassug{Reviewer C, esop, would have liked a discussion of (a) what could be done on a machine with a smaller set of capabilities, e.g., no local capability (we could probably not have done anything as the stack pointer could be stored on the heap and reused at unintended times.) and (b) what could be done with a larger/stronger set of available capabilities (we could include a discussion of linear capabilities).}
\itoplassug{Reviewer C, esop, asks whether our attacker model is reasonable (see tex comment). Maybe we should include a short discussion of the attacker model.}
% p.18 "they are linked with an adversary, adv, that is allowed to allocate memory but has no other capabilities".  Is this a reasonable model of the adversary?  (Genuinely naive question.)
\itoplassug{Should we include a discussion of how this ``scales'' to other things like a multi-core setting or if we needed tail calls?}
% * Also, the local capabilities seem related to Boyland's "Borrowing"
%   e.g. that is now supported in Rust amongst other research languages
%   (Pony etc) also mentioned in the ECOOP paper.
% * the combination of E and L capabilities seems very interesting
%   indeed.  I wonder if e.g. they could be used to support strong
%   models of encapsulation e.g. like ownership types. (see eg LNCS
%   7850)
\noindent\textbf{Calling convention}\\
\emph{Formulating control flow correctness} While we claim that our calling
convention enforces control-flow correctness, we do not prove a general theorem
that shows this because it is not clear what such a theorem should look like.
Formulations in terms of a control-flow graph, like the one by \citet{abadi_control-flow_2005}, do not take into account temporal properties \itoplas{because they only take into account what capabilities can be available and not when they are used.
Example~\texttt{g1} relies on such temporal properties when returning from the higher-order call backs.}
In fact, our examples show that our logical relation implies a stronger form of
control-flow correctness than such formulations, although this is not made very
explicit. As future work, we consider looking at a more explicit and useful way
to formalize control-flow correctness. The idea would be to define a variant
of our capability machine with call and return instructions and
well-bracketed control flow built-in to the operational semantics, and then
prove that compiling such code to our machine using our calling convention is
fully abstract~\citep{abadi_protection_1998}.
\itoplassug{Maybe elaborate on why full-abstractness and not weaker properties like ``robust safety''.}

@@

A second important remark we want to make is that the need for stack clearing in our calling convention is an instance of a general caveat when using CHERI's local capabilities as a restricted form of capability revocation.
Consider how our use of local capabilities can be interpreted as temporarily delegating the stack and return capability to callees and then revoking the granted authority after the callee returns.
From this perspective, local capabilities are a general feature enabling this temporary delegation of authority for the duration of an invocation and this is also how they are described by the authors~\citep{Watson2015Cheri}.
However, our requirement for stack clearing on boundary crossings is also general.
Revoking authority that was granted temporarily using local capabilities requires clearing all memory for which the invokee had write-local authority (or at least erasing all local capabilities from that memory).
Without micro-architectural support for efficiently clearing large ranges of memory, local capabilities can only be used for revocation in scenarios where the duration of a revocation is unimportant or the adversary only has write-local access to small amounts of memory.
In future work, we are planning to investigate an alternative notion of \textit{linear} capabilities that offer a limited form of revocation without such limitations.

@@

\itoplassug{We could also include a brief discussion of concurrency and the fact that this probably wouldn't scale to it (but well-bracktedness is not the same in that case anyway.) }
\itoplassug{Reviewer C, esop, wants to know what happens when there are several stack blocks (see tex for comment). We could here emphasize that while we talk about a single stack there could be multiple local read/write-local/execute-capabilities. If we get the stack from an untrusted source, which is the case in the awkward example, then we will use anything that \emph{looks like} the stack as the stack. However, due to the operational semantics for local capabilities all other ``stack blocks'' would have to be stored on the stack or in the register file. As the calling convention dictates that we clear these, the only place the alternative stack blocks will not be cleared is on a part of the stack that we do not have access to. The return capability would basically have to encapsulate this part of memory (perhaps indirectly) which means that from the moment we have been called, we will make sure there is only one stack available until our program returns from the initial call to it.}
%p.10 "to prevent this from happening, we need to make sure the stack capability carries RWLX authority since the system wide assumption then tells us that the adversary cannot have global capabilities to our local stack".  It feels like the "L" capabitility is being overloaded to mean not just "this memory block has write-local authority and can contain local capabilities", but also to mean "this memory block is my stack and no other block is".  What happens when there are several stack blocks, e.g. one per thread?
\emph{Single stack} A single stack is a good choice for the simple capability machine presented here because it works well with higher-order functions.
An alternative to a single stack would be to have a separate stack per component.
The trouble with this approach is that, with multiple stacks and local stack pointers, it is not clear how components would retrieve their stack pointer upon invocation without compromising safety.
A safe approach could be to have stack pointers stored by a central, trusted stack management component, but it is not clear how that could scale to large numbers of separate components.
Handling large numbers of components is a requirement if we want to use capability machines to enforce encapsulation of, for example, every object in an object-oriented program or every closure in a functional program.

\\\\
\begin{toplas}
  \noindent \textbf{Capability machine formalization}\\
  \emph{Simplifications} Our capability machine formalization assumes unbounded integers and an infinite address space.
  Further, it has a much simpler instruction set than that of a full ISA.
  By making these simplifications, we avoid tedious details but end up with a less realistic machine.
  However, the intend of this work is to gain ground in the formal work necessary to prove properties about low-level assembly programs on a capability machine not to apply it to a full fledged capability machine.

@@

If we want to scale this approach even further, we would like to reason at a higher level of abstraction namely at the level of a high-level language.
That is, we would rather construct a program logic for a high-level programming language, so we can reason about the programs in the language we actually write them in and at the same time get guarantees about the compiled program.
To achieve this, we would have to construct a secure compilation \citep{patrignani_secure_compilation_2019} that preserves the security abstraction of the high-level language.
The experiences made about logical relations for capability machines here has been used by \citet{skorstengaard_stktokens_2019} to make advances in the direction of building a secure compilation targeting capability machines.
They present a calling convention based on linear capabilities rather than local capabilities, and they provide a fully-abstract compilation that can be used as the last phase in a secure compilation.
The full-abstraction proof uses a logical relation.
\end{toplas}
\itoplassug{We could further add that this is not the level of abstraction that
  one would like to reason. One would like to reason about the programs we
  actually write - that is programs written in high-level languages. If we had a
  fully-abstract compiler from a nice high-level language to this low-level
  machine, then we could reason about the program in the high-level language and
  then because the compiler is fully-abstract, we would retain all the nice
  properties from the high-level language. How does this work fit in to this
  picture? The calling convention ensures two properties that we expect our nice
  high-level languages to have, namely well-bracketedness and local-state
  encapsulation. If we want to do fully-abstract compilation, then we need some
  way to guarantee this after compilation. Further, logical relations are often
  used in full-abstractness proofs. This work expands our knowledge about
  logical relations should be defined for low-languages and in particular
  capability machines. (maybe also mention that enforcement necessary for
  fully-abstract compilation. Could be capability, but it could also be
  something else - we just need something which we do not have now.) }

>>>>

\section{Discussion}
\label{sec:discussion}
\itoplassug{Reviewer C, esop, would have liked a discussion of (a) what could be done on a machine with a smaller set of capabilities, e.g., no local capability (we could probably not have done anything as the stack pointer could be stored on the heap and reused at unintended times.) and (b) what could be done with a larger/stronger set of available capabilities (we could include a discussion of linear capabilities).}
\itoplassug{Reviewer C, esop, asks whether our attacker model is reasonable (see tex comment). Maybe we should include a short discussion of the attacker model.}
% p.18 "they are linked with an adversary, adv, that is allowed to allocate memory but has no other capabilities".  Is this a reasonable model of the adversary?  (Genuinely naive question.)
\itoplassug{Should we include a discussion of how this ``scales'' to other things like a multi-core setting or if we needed tail calls?}
% * Also, the local capabilities seem related to Boyland's "Borrowing"
%   e.g. that is now supported in Rust amongst other research languages
%   (Pony etc) also mentioned in the ECOOP paper.
% * the combination of E and L capabilities seems very interesting
%   indeed.  I wonder if e.g. they could be used to support strong
%   models of encapsulation e.g. like ownership types. (see eg LNCS
%   7850)
\noindent\textbf{Calling convention}\\
\emph{Formulating control flow correctness} While we claim that our calling convention enforces control-flow correctness, we do not prove a general theorem that shows this because it is not clear what such a theorem should look like.
Formulations in terms of a control-flow graph, like the one by \citet{abadi_control-flow_2005}, do not take into account temporal properties \itoplas{because they only take into account what capabilities can be available and not when they are used.
  Example~\texttt{g1} relies on such temporal properties when returning from the higher-order call backs.}
In fact, our examples show that our logical relation implies a stronger form of control-flow correctness than such formulations, although this is not made very explicit.
In later work, \citet{skorstengaard_stktokens_2019} provide a more explicit and useful definition of control-flow correctness and local state encapsulation.
The idea of their definition is to define a variant of a simple capability machine with a built-in stack as well as call and return instructions.
The call and return instructions use the built-in stack which gives well-bracketed control and local state encapsulation by construction.
They prove full abstraction~\citep{abadi_protection_1998} for a compilation from the capability machine with the built-in stack to the one without which means that the programs running on the capability machine without a built-in stack behaves as though there was a built-in stack.
Their full abstraction proof uses a logical relation, but it differs in a number of ways.
First, their logical relation is a binary cross language logical relation tailored specifically to prove the full abstraction result.
This means that it could not be used to prove program correctness results similar to ours.
Their calling convention is based on something called linear capabilities that offer a different kind of limited revocation.
\itoplassug{Maybe elaborate on why full-abstractness and not weaker properties like ``robust safety''.}

@@

A second important remark we want to make is that the need for stack clearing in our calling convention is an instance of a general caveat when using CHERI's local capabilities as a restricted form of capability revocation.
Consider how our use of local capabilities can be interpreted as temporarily delegating the stack and return capability to callees and then revoking the granted authority after the callee returns.
From this perspective, local capabilities are a general feature enabling this temporary delegation of authority for the duration of an invocation and this is also how they are described by the authors~\citep{Watson2015Cheri}.
However, our requirement for stack clearing on boundary crossings is also general.
Revoking authority that was granted temporarily using local capabilities requires clearing all memory for which the invokee had write-local authority (or at least erasing all local capabilities from that memory).
Without micro-architectural support for efficiently clearing large ranges of memory, local capabilities can only be used for revocation in scenarios where the duration of a revocation is unimportant or the adversary only has write-local access to small amounts of memory.

@@

\itoplassug{We could also include a brief discussion of concurrency and the fact that this probably wouldn't scale to it (but well-bracktedness is not the same in that case anyway.) }
\itoplassug{Reviewer C, esop, wants to know what happens when there are several stack blocks (see tex for comment). We could here emphasize that while we talk about a single stack there could be multiple local read/write-local/execute-capabilities. If we get the stack from an untrusted source, which is the case in the awkward example, then we will use anything that \emph{looks like} the stack as the stack. However, due to the operational semantics for local capabilities all other ``stack blocks'' would have to be stored on the stack or in the register file. As the calling convention dictates that we clear these, the only place the alternative stack blocks will not be cleared is on a part of the stack that we do not have access to. The return capability would basically have to encapsulate this part of memory (perhaps indirectly) which means that from the moment we have been called, we will make sure there is only one stack available until our program returns from the initial call to it.}
%p.10 "to prevent this from happening, we need to make sure the stack capability carries RWLX authority since the system wide assumption then tells us that the adversary cannot have global capabilities to our local stack".  It feels like the "L" capabitility is being overloaded to mean not just "this memory block has write-local authority and can contain local capabilities", but also to mean "this memory block is my stack and no other block is".  What happens when there are several stack blocks, e.g. one per thread?
\emph{Single stack} A single stack is a good choice for the simple capability machine presented here because it works well with higher-order functions.
An alternative to a single stack would be to have a separate stack per component.
The trouble with this approach is that, with multiple stacks and local stack pointers, it is not clear how components would retrieve their stack pointer upon invocation without compromising safety.
A safe approach could be to have stack pointers stored by a central, trusted stack management component, but it is not clear how that could scale to large numbers of separate components.
Handling large numbers of components is a requirement if we want to use capability machines to enforce encapsulation of, for example, every object in an object-oriented program or every closure in a functional program.
\\\\
\begin{toplas}
  \noindent \textbf{Capability machine formalization}\\
  \emph{Simplifications} Our capability machine formalization assumes unbounded integers and an infinite address space.
  Further, it has a much simpler instruction set than that of a full ISA.
  By making these simplifications, we avoid tedious details but end up with a less realistic machine.
  However, the intend of this work is to gain ground in the formal work necessary to prove properties about low-level assembly programs on a capability machine not to apply it to a full fledged capability machine.

@@

If we want to scale this approach even further, we would like to reason at a higher level of abstraction namely at the level of a high-level language.
That is, we would rather construct a program logic for a high-level programming language, so we can reason about the programs in the language we actually write them in and at the same time get guarantees about the compiled program.
To achieve this, we would have to construct a secure compilation \citep{patrignani_secure_compilation_2019} that preserves the security abstraction of the high-level language.
\end{toplas}
\itoplassug{We could further add that this is not the level of abstraction that
  one would like to reason. One would like to reason about the programs we
  actually write - that is programs written in high-level languages. If we had a
  fully-abstract compiler from a nice high-level language to this low-level
  machine, then we could reason about the program in the high-level language and
  then because the compiler is fully-abstract, we would retain all the nice
  properties from the high-level language. How does this work fit in to this
  picture? The calling convention ensures two properties that we expect our nice
  high-level languages to have, namely well-bracketedness and local-state
  encapsulation. If we want to do fully-abstract compilation, then we need some
  way to guarantee this after compilation. Further, logical relations are often
  used in full-abstractness proofs. This work expands our knowledge about
  logical relations should be defined for low-languages and in particular
  capability machines. (maybe also mention that enforcement necessary for
  fully-abstract compilation. Could be capability, but it could also be
  something else - we just need something which we do not have now.) }
